---
title: "csdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{csdata}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(csdata)
library(data.table)
library(magrittr)
```

`csdata` provides structural reference data for Norway: location hierarchies, population counts, and the conventions used across the csverse format.

See https://niphr.github.io/csdata/reference/index.html for an overview of all available datasets and functions.

## What csdata is for

`csdata` holds the reference tables that Norwegian surveillance work needs again and again: which places exist and how they nest inside each other, how place boundaries were redistricted from 2006 to 2024, and how many people live in each place by age and sex. Every table ships inside the package, so nothing is downloaded when you run it. The point is that each analysis stops carrying its own copy of the same lookups.

Thirteen objects are exported. These five do most of the work.

| Function | Use it when you need |
|---|---|
| `nor_locations_names()` | The list of places, and which level each place sits at. |
| `nor_locations_hierarchy_from_to(from, to)` | To map one level onto another, such as every municipality to its county. |
| `nor_locations_redistricting()` | Weights to move data recorded under older borders onto the 2024 borders. |
| `nor_population_by_age_cats(cats)` | Denominators: population in the age bands you choose. |
| `nor_population_by_sex_age_cats(cats)` | The same denominators, split by sex. |

All five take a `border` argument. `2024` is the only value they accept; any other value stops with an error.

## Two current limitations

**BA-regions are missing from the hierarchy table.** `nor_locations_names()` lists 159 BA-regions, and `nor_locations_hierarchy_from_to()` accepts `"baregion"`. But the bundled hierarchy table carries no BA-region codes at all, so every `from`/`to` pair naming `"baregion"` returns an empty table instead of an error.

```{r}
nrow(csdata::nor_locations_hierarchy_from_to(from = "municip", to = "baregion"))
nrow(csdata::nor_locations_names()[granularity_geo == "baregion"])
```

**One location code is used twice.** Two different laboratories share the code `lab_nor084467`.

```{r}
loc <- csdata::nor_locations_names()
loc[location_code == "lab_nor084467", .(location_code, location_name)]
```

This matters when you pass `nor_locations_names()` as `location_reference`, because the lookup is a join on `location_code`. The duplicated code matches twice, so you get more values back than you put in.

```{r}
csdata::location_code_to_granularity_geo(
  c("nation_nor", "county_nor03", "lab_nor084467"),
  location_reference = loc
)
```

`add_granularity_geo_to_data_set()` then fails, because it tries to write those four values into a three-row table.

```{r, error = TRUE}
x <- data.table(location_code = c("nation_nor", "county_nor03", "lab_nor084467"))
csdata::add_granularity_geo_to_data_set(x, location_reference = loc)
```

Leave `location_reference` unset to avoid both problems. The granularity is then read from the code prefix, one value per input.

## Where csdata sits, and what to read next

`csdata` depends on no other cs package, so you can install it on its own. `cstidy` imports it. `csmaps` uses the same `location_code` values for the 2024 borders, so its map polygons join straight onto csdata tables.

Two vignettes go deeper:

- `vignette("locations_norway", package = "csdata")` lists every location code with its name.
- `vignette("population_norway", package = "csdata")` shows population counts by location and calendar year.

The rest of this page gives the csverse coding rules for locations, ages and sex.

## Location

Valid locations and location types are available in `csdata::nor_locations_names()`. Entries with uncommon or internal use are shown with a strikethrough.

```{r echo=FALSE, results='asis'}
d <- csdata::nor_locations_names()[, .(
  n = .N,
  location_code = location_code[1],
  location_name = location_name[1],
  location_name_description_nb = location_name_description_nb[1],
  location_name_file_nb_utf = location_name_file_nb_utf[1],
  location_name_file_nb_ascii = location_name_file_nb_ascii[1]
),
by = .(granularity_geo)
]

gt::gt(d) %>%
  gt::tab_options(
    table.width = "1500px"
  ) %>%
  gt::tab_header(title = "Valid locations and location types in the csverse format") %>%
  gt::cols_label(
    granularity_geo = "Geo (Granularity)",
    n = "N"
  ) %>%
  # gt::cols_width(
  #   granularity_time ~ "20%",
  #   class ~ "15%",
  #   fn ~ "20%",
  #   example ~ "55%"
  # ) %>%
  gt::tab_spanner(
    label = "Examples",
    columns = c(location_code, location_name, location_name_description_nb, location_name_file_nb_utf, location_name_file_nb_ascii)
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_code**: Used a) **inside datasets** and b) in data **file names** for transfer of data/results between analytic systems. All values are unique."),
    locations = gt::cells_column_labels(
      columns = location_code
    )
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_name**: Used (rarely) **inside results** (figures, tables, documents). Can be confusing as some names are duplicated. Its rare usage is demarcated by a line through the text."),
    locations = gt::cells_column_labels(
      columns = location_name
    )
  ) %>%
  gt::tab_style(
    style = list(
      gt::cell_text(decorate = "line-through")
    ),
    locations = gt::cells_body(
      columns = location_name,
      rows = gt::everything()
    )
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_name_description_nb**: Used (frequently) **inside results** (figures, tables, documents). All values are unique."),
    locations = gt::cells_column_labels(
      columns = location_name_description_nb
    )
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_name_file_nb_utf**: Used (frequently) in the **file names** for results (figures, tables, documents). All values are unique."),
    locations = gt::cells_column_labels(
      columns = location_name_file_nb_utf
    )
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_name_file_nb_ascii**: Used (rarely) in the **file names** for results (figures, tables, documents). Used if file systems have problems with the Norwegian letters æøå. All values are unique."),
    locations = gt::cells_column_labels(
      columns = location_name_file_nb_ascii
    )
  ) %>%
  gt::tab_footnote(
    footnote = "Bo- og arbeidsmarkedsregioner. Housing and labor market regions.",
    locations = gt::cells_body(
      columns = granularity_geo,
      rows = granularity_geo == "baregion"
    )
  ) %>%
  gt::tab_footnote(
    footnote = "Landsdeler/riskdeler. Geographical regions.",
    locations = gt::cells_body(
      columns = granularity_geo,
      rows = granularity_geo == "georegion"
    )
  ) %>%
  gt::tab_footnote(
    footnote = "Mattilsynet-regioner. Food authority regions.",
    locations = gt::cells_body(
      columns = granularity_geo,
      rows = granularity_geo == "mtregion"
    )
  )
```

## Ages

Ages must be coded as characters and must always contain three digits. For age ranges, join the two ages with an underscore (e.g. `005_010`).

Use `085p` rather than `>=085` or `85+`; this keeps conversion from long to wide format straightforward.

```{r echo=FALSE, results='asis'}
d <- rbind(
  data.frame(
    value = "\"000\"",
    class = "character",
    definition = "One year age group (0 year olds)"
  ),
  data.frame(
    value = "\"079\"",
    class = "character",
    definition = "One year age group(79 year olds)"
  ),
  data.frame(
    value = "\"000_004\"",
    class = "character",
    definition = "Age span of 0-4 year olds"
  ),
  data.frame(
    value = "\"065p\"",
    class = "character",
    definition = "Age span of >=65 year olds"
  ),
  data.frame(
    value = "\"missing\"",
    class = "character",
    definition = "Missing/unknown"
  ),
  data.frame(
    value = "\"total\"",
    class = "character",
    definition = "Everyone"
  )
)

gt::gt(d) %>%
  gt::tab_header(title = "Valid ages in the csverse format") %>%
  gt::cols_label(
    value = "Value",
    definition = "Definition"
  )
```

This format keeps data sorted correctly and produces valid variable names when pivoted to wide format.

Missing ages should be coded as `"missing"`.

## Sex

Sex must be coded as a character.

```{r echo=FALSE, results='asis'}
d <- rbind(
  data.frame(
    value = "\"male\"",
    class = "character",
    definition = "Male"
  ),
  data.frame(
    value = "\"female\"",
    class = "character",
    definition = "Female"
  ),
  data.frame(
    value = "\"missing\"",
    class = "character",
    definition = "Missing/unknown"
  ),
  data.frame(
    value = "\"total\"",
    class = "character",
    definition = "Everyone"
  )
)

gt::gt(d) %>%
  gt::tab_header(title = "Valid sexes in the csverse format") %>%
  gt::cols_label(
    value = "Value",
    definition = "Definition"
  )
```

Missing sex should be coded as `"missing"`.
